Data Specialization for a Staged Lambda Calculus

### 2015-06-23 paper/semantics.tex
editted statics

We express two-stage programs as terms in \lang, a typed, modal lambda calculus
with products, sums, and isorecursive types. Its type system has three distinct
\emph{worlds} \bbonep, \bbonem, and \bbtwo, which respectively classify
\emph{ground first-stage} computations (\bbonep); 
first-stage computations containing a \emph{mix} of code from each stage
(\bbonem); 
and second-stage computations (\bbtwo).

\subsection{Statics}

@@

The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
$e$ has type $A$ at world $w$, in the context $\Gamma$. (Variables in the
context are also annotated with the world at which they live.)

All three worlds contain fixpoints and unit, product, sum, function, and
recursive types defined in the usual fashion. For example, given two terms
$\typesone{e_1}A$ and $\typesone{e_2}B$ at world \bbonem, one can form the pair
$\typesone{\texttt{<}e_1,e_2\texttt{>}}{A\times B}$ also at world \bbonem.

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers. $\fut A$ is a type in world \bbonem\
which classifies second-stage computations of type $A$. Given a term $e$ of type
$A$ at world \bbtwo, $\next{e}$ has type $\fut A$ at \bbonem. This essentially
encapsulates $e$ as a computation to be evaluated in the future---first-stage
computations can shuffle it around, but not use its result.  The only way to use
a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.

%This is why well-typed terms cannot have any information flow from stage
%\bbtwo\ to stage \bbone, and what makes splitting possible.

@@

\subsection{(Small Step) Dynamics}
We provide a small-step semantics which,
given a \lang\ term $e$, evaluates its first stage subterms
either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
or until all of $e$'s first stage subterms have been evaluated away
(when $e$ is at world \bbtwo).
Because first- and second-stage code are
interleaved in \lang\ terms, this first pass of
evaluation requires some care.

>>>>

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
The type system of \lang\ (\ref{sec:statics}) classifies both the stage and type of a 
computation, and is sufficient to show that the staged dynamics of \lang\ (\ref{sec:dynamics})
do not get stuck.

\subsection{Statics}
\label{sec:statics}

@@

The typing judgment $\typeswor e A$, defined in \ref{fig:statics}, means that
$e$ has type $A$ at world $w$, in the context $\Gamma$. 
Here, a \emph{world} is a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonep\ for \emph{ground} first-stage computations, and 
\bbonem\ for first-stage computations containing a \emph{mix} of code from each stage. 
The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
and will be discussed in \ref{sec:needGround}.

All three worlds contain unit, product, and function types defined in the usual fashion;
sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonep.
For example, given a term $e$ of type $A$ at world \bbtwo, 
$\next{e}$ has type $\fut A$ at \bbonem. This essentially
encapsulates $e$ as a computation to be evaluated in the future---first-stage
computations can shuffle it around, but not use its result.  The only way to use
a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.

@@

\subsection{(Small Step) Dynamics}
\label{sec:dynamics}

We provide a small-step semantics which,
given a \lang\ term $e$, evaluates its first stage subterms
either until $e$ is a value (if $e$ is at world \bbonep\ or \bbonem)
or until all of $e$'s first stage subterms have been evaluated away
(when $e$ is at world \bbtwo).
Because first- and second-stage code are
interleaved in \lang\ terms, this first pass of
evaluation requires some care.
